Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-11-25 rnni_convexity.tex
pass to caterpillar sort + rm Lemma 1 proof details (some)

@@

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
\todo{Geometry summary}
Specifically, we...

We also study the problem of computing distances between trees in this graph.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we specifically focus on the properties that have already been established for $\nni$.
Surprisingly, our results suggest that the complexity of computing distances in the two graphs is likely to be different.
\todo{Algorithms summary}
Specifically, we...
\end{abstract}

@@

Now we are ready to define the tree space which is the subject of study in this paper, the $\rnni$ graph.

@@

@@

We start this section with a discussion of the algorithm proposed in \autocite{Gavryushkin2018-ol} to compute the exact $\rnni$ distance between trees by pre-computing the $\rnni$ graph.
Then we introduce two algorithms, $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
These algorithms will become an important ingredient to establish several geometric properties of the $\rnni$ graph in Section~\ref{section:geometry}.
Finally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree (see Section~\ref{section:alg_mdtree} for details).
While this algorithm is interesting on its own (e.g.\ for simulation studies), we will again apply the algorithm to study the geometry of the $\rnni$ space.

For comparing approximate distances computed by $\findpath$ (Section~\ref{section:alg_findpath}) with exact distances and for supporting Conjecture~\ref{conjecture:cluster_theorem}, we aim to compute exact distances in $\rnni$ for small trees.
Therefore we will use the algorithm for computing the $\rnni$ graph that is given in \textcite[Section 3.3]{Gavryushkin2018-ol}.
\todo{Do we make use of tanglegrams? - LC: No}
\todo{GitHub link?}
Once the graph is computed one can use common algorithms for computing shortest paths~\autocite{Floyd1962-ew,Dijkstra1959-ph}.
However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances.
Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we are just able to use the $\rnni$ graph on up to six taxa for our analyses.
We will get into further details when we discuss particular problems in Section~\ref{section:alg_findpath} and~\ref{section:cluster_theorem}.

@@

Within this section we are presenting $\findpath$, an efficient algorithm for computing paths, and therefore approximating distances, between any given pair of trees in $\rnni$.
As we will see in Section~\ref{section:geometry}, it is useful for gaining insights into geometric properties of the $\rnni$ graph.

@@

We are now ready to explain how $\findpath$ (Algorithm~\ref{alg:find_path}) computes a path between two input trees $T$ and $R$.
Let $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
Then $C_i$ is induced by the node of rank $i$ for $i = 1, \ldots, n-2$.
$\findpath$ computes a path which initially consists of $T_0 = T$ only and is extended stepwise as described in the following.

Step $0$.
$p = [T_0]$.
Consider the node in $T_0$ that induces the smallest cluster containing all elements of $C_1$.
This node is denoted by $\mrca_{T_0}(C_1)$ (short for \emph{most recent common ancestor}).
By performing an $\rnni$ move that decreases the rank of $\mrca_{T_0}(C_1)$, we receive a tree $T_1$ that is then appended to $p$.
Now we repeatedly perform $\rnni$ moves that move the most recent common ancestor of $C_1$ down in order to extend $p$ to $p = [T_0, \ldots, T_l]$ for some $l$ such that $C_1$ is induced by the internal node of rank one in $T_l$.

@@

\todo{$p = p+T'$: Do we really want to be updating this in every while loop? - LC: Yes, I guess so}
\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
\STATE $T' := T$
\STATE $p := [T']$
\FOR {$k = 1, \dots, n-2$}
\STATE Let $C_k$ be the cluster induced by the node with rank $k$ in $R$ \label{alg:find_path:line:cluster}
\WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
\STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
\STATE $p = p+T'$
\ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

The following lemma shows that $\findpath$ gives a path between the two input trees, indeed, by proving that there is always an $\rnni$ move possible as described in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.

@@

% It is important to notice that $C_k$ either is the disjoint union $C_k = C_i \dot\cup C_j$ for $i,j < k$ or $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
%
% Let us at first assume that $C_k = C_i \dot\cup C_j$ for some $i,j < k$.
% The tree $T$ looks like the tree depicted in the left of Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
% Since we know that all clusters $C_1,\ ldots, C_{k-1}$ are induced by some nodes in $T$, each of these clusters must be a subset of the taxon set of one of the three subtrees $T_w^1,T_w^2$, or $T_v$.
% In particular, there can't be elements of the same cluster $C_m$ in more than one of these subtrees.
% This and the fact that $v = \mrca_{T}(C_i \dot\cup C_j)$ yields that either $C_i$ or $C_j$ is subset of the taxon set of $T_v$ and the other cluster is subset of the taxon set of either $T_w^1$ or $T_w^2$.
%
% Without loss of generality we assume that $C_i$ is included in $T_v$.
% Otherwise, change the indices $i$ and $j$.
% Then $C_i$ is a subset of the taxon set of either $T_w^1$ or $T_w^2$.
% Therefore, just one of the two $\nni$ moves on $(v,w)$, as depicted in Figure~\ref{fig:mrca_move}, moves the most recent common ancestor of $C_k = C_i \dot \cup C_j$ down.
%
% Now we consider the case that $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
% For the same reason as above, $C_i$ is in subset of the taxon set of one of the subtrees $T_w^1, T_w^2$ or $T_v$ of $\hat T$ as depicted in Figure~\ref{fig:mrca_move}.
% Similarly $x$ is taxon of one of these three subtrees in a way that either $C_i$ or $x$ is included in $T_v$, because $v = \mrca_{T'}(C_i \dot \cup \{x\})$.
% So there is just one of the two subtrees $T_w^1$ and $T_w^2$ that neither includes $C_j$ nor $x$.
% As in the previous case, just one of the two $\nni$ moves on $(v,w)$ as illustrated in Figure~\ref{fig:mrca_move} decreases the rank of $\mrca_{T}(C_i \dot \cup \{x\})$.
%
% We can conclude that in any case there is exactly one $\rnni$ move possible on $T$ that decreases the rank of $\mrca_{T}(C_k)$ in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.

\begin{figure}[H]
\centering
\includegraphics[width=0.5\textwidth]{mrca_move}
\vspace{12pt}
\caption{Two $\nni$ moves that are possible on the edge $(v,w)$ of the tree $T'$ with subtrees $T_v, T_{w_1}, T_{w_2}$ as described in the proof of Lemma~\ref{lemma:mrca_move}.}
\label{fig:mrca_move}
\end{figure}

\end{proof}

Note that the uniqueness of the $\rnni$ move in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} implies that the algorithm is deterministic.
The running time of the algorithm is quadratic in the number of taxa.

>>>>

\sloppy

@@

\begin{abstract}
In this paper we study the graph of ranked phylogenetic trees where the adjacency relation is given by a local rearrangement of the tree structure.
Our work is motivated by tree inference algorithms, such as maximum likelihood and Markov Chain Monte Carlo methods, where the geometry of the search space plays a central role for efficiency and practicality of the optimisation.
We hence focus on understanding the geometry of the space (graph) of ranked trees, the so-called ranked nearest neighbour interchange ($\rnni$) graph.
%TODO
Specifically, we...

We also study the problem of computing distances between trees in this graph.
Since the $\rnni$ graph is a generalisation of the classical nearest neighbour interchange ($\nni$) graph to ranked phylogenetic trees, we specifically focus on the properties that have already been established for $\nni$.
Surprisingly, our results suggest that the complexity of computing distances in the two graphs is likely to be different.
%TODO
Specifically, we...
\end{abstract}

@@

Now we are ready to introduce the tree space which is the subject of study in this paper, the $\rnni$ graph.

@@

\todo{Paper structure}

@@

In this section we introduce three algorithms, $\findpath$ (Section~\ref{section:alg_findpath}), $\csort$ (Section~\ref{section:alg_csort}), and $\mdtree$ (Section~\ref{section:alg_mdtree}).
$\findpath$ computes a $\rnni$ path between trees.
$\csort$ computes a shortest $\rnni$ path between caterpillar trees (also known as ladder trees, see later in this section for accurate definitions).
$\mdtree$ maximises the dissimilarity to find a remote tree given an arbitrary $\rnni$ tree.
While these algorithms are interesting on their own (e.g.\ in simulation studies), they will also form an important ingredient to study the geometry of the $\rnni$ space later in this paper.

The algorithm $\findpath$ can be used to approximate the $\rnni$ distance.
We will study the accuracy of this approximation by computing the exact $\rnni$ distance for small trees.
Specifically, we will use the algorithm for computing the complete $\rnni$ graph that is given in \autocite[their Section 3.3]{Gavryushkin2018-ol}.
\todo{GitHub link?}
After pre-computing the graph, we use the standard algorithms~\autocite{Floyd1962-ew,Dijkstra1959-ph} for computing shortest paths.
However, the large number of vertices in $\rnni$ is an obstacle when it comes to computing distances for larger trees.
Since there are $\frac{n!(n-1)!}{2^{n-1}}$ vertices in the $\rnni$ graph \autocite{Gavryushkin2018-ol}, we compute the $\rnni$ graph on up to six taxa for our analyses.
This computational result will also be used later in the paper to support Conjecture~\ref{conjecture:cluster_theorem}, a modification of Conjecture~9 in \autocite{Gavryushkin2018-ol}, which we prove to be false in this paper.

@@

In this section we present $\findpath$, an efficient algorithm for computing paths, and therefore approximating distances, between trees in $\rnni$.

@@

Let $T$ and $R$ be arbitrary trees and $[C_1, \ldots, C_{n-2}]$ be the cluster representation of $R$ (the ``destination'' tree).
Then $C_i$ is induced by the node of rank $i$ for $i = 1, \ldots, n-2$.
$\findpath$ computes a path by iteratively extending a sequence $p$ of trees starting from $T$.
The algorithm terminates when $p$ is an $\rnni$ path from $T$ to $R$.

\todo{Alex: go through these steps and proof of L1}
Step $0$.
$p = [T_0]$.
Consider the node in $T_0$ that induces the smallest cluster containing all elements of $C_1$.
This node is denoted by $\mrca_{T_0}(C_1)$ (short for \emph{most recent common ancestor}).
By performing an $\rnni$ move that decreases the rank of $\mrca_{T_0}(C_1)$, we receive a tree $T_1$ that is then appended to $p$.
Now we repeatedly perform $\rnni$ moves that move the most recent common ancestor of $C_1$ down in order to extend $p$ to $p = [T_0, \ldots, T_l]$ for some $l$ such that $C_1$ is induced by the internal node of rank one in $T_l$.

@@

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
\STATE $T' := T$
\STATE $p := [T']$
\FOR {$k = 1, \dots, n-2$}
\STATE Let $C_k$ be the cluster induced by the node with rank $k$ in $R$ \label{alg:find_path:line:cluster}
\WHILE {$\rank_{T'}(\mrca_{T'}(C_k))>k$}
\STATE Update $T'$: Decrease $\rank_{T'}(\mrca_{T'}(C_k))$ by an $\rnni$ move \label{alg:findpath:line:move_set_down}
\STATE $p = p+T'$
\ENDWHILE
\ENDFOR
\RETURN $p$
\end{algorithmic}
\end{algorithm}

The following lemma shows that $\findpath$ is correct, that is, the algorithm returns a path between the two input trees.
To prove this we have to establish that there always exists a unique $\rnni$ move required in Line~\ref{alg:findpath:line:move_set_down} of the algorithm.

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.5\textwidth]{mrca_move}
\vspace{12pt}
\caption{Two $\nni$ moves that are possible on the edge $(v,w)$ of the tree $T'$ with subtrees $T_v, T_{w_1}, T_{w_2}$ as described in the proof of Lemma~\ref{lemma:mrca_move}.}
\label{fig:mrca_move}
\end{figure}
\end{proof}

Note that the uniqueness of the $\rnni$ move in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} also implies that the algorithm is deterministic.
The running time of the algorithm is quadratic in the number of taxa.
